---
title: "Chapter 2 Solutions"
author: "Cuong Duong"
date: "Last updated: `r format(.POSIXct(Sys.time(), tz = 'UTC'), '%Y-%m-%d %H:%M:%S %Z')`"
output:
  html_notebook:
    default
---

```{r, eval=TRUE, results='hide'}
library(cmdstanr)
library(ggplot2)
library(data.table)
library(magrittr)

options(scipen = 999)
theme_set(theme_minimal())
```

### 2E1

#### Answer

Both `(2)` and `(4)` correspond to the probability of rain on Monday. `(4)` is an expanded version of the conditional probability expression in `(2)`.

### 2E2

#### Answer

`(3)` The probability that it is Monday, given that it is raining.

### 2E3

#### Answer

Both `(1)` and `(4)` correspond to the probability that is Monday given that it is raining. `(4)` re-writes `(1)` using Bayes' Theorem.

### 2E4

#### Answer

"The probability of water is 0.7" refers to our uncertainty of the proportion of the Earth's surface that is water. In this context, we assume we will never know the true water proportion (not true in reality, but could be the case if the proportion of land and water continually change over time). Hence to us, the water proportion is a distribution of possible values, and we update this distribution in light of new relevant data - the globe tosses.

### 2M1

#### Answer

```{r}
# Uniform prior
prior_uniform <- function(p) {
  p
}

calculate_likelihood <- function(obs, p) {
  values <- ifelse(obs == 'W', 1, 0)
  dbinom(x = sum(values), size = length(values), p = p)
}

posterior_grid_approximate <- function(prior_function, obs, n_points) {
  params <- seq(0, 1, length.out = n_points)
  prior <- prior_function(params)
  likelihood <- calculate_likelihood(obs, params)
  product <- prior * likelihood
  probs <- product / sum(product)
  
  data.table(p = params, prob = probs)
}

plot_posterior_grid <- function(df) {
  ggplot(df, aes(x = p, y = prob)) +
    geom_point()
}
```

(1)

```{r}
posterior_grid_approximate(prior_uniform, c('W', 'W', 'W'), n_points = 50) %>% 
  plot_posterior_grid()
```
(2)

```{r}
posterior_grid_approximate(prior_uniform, c('W', 'W', 'W', 'L'), n_points = 50) %>% 
  plot_posterior_grid()
```

(3)

```{r}
posterior_grid_approximate(prior_uniform, c('L', 'W', 'W', 'L', 'W', 'W', 'W'), n_points = 50) %>% 
  plot_posterior_grid()
```
### 2M2

#### Answer

```{r}
prior_step <- function(p) {
  ifelse(p < 0.5, 0, 1 / (1 - 0.5))
}
```

(1)

```{r}
posterior_grid_approximate(prior_step, c('W', 'W', 'W'), n_points = 50) %>% 
  plot_posterior_grid()
```

(2)

```{r}
posterior_grid_approximate(prior_step, c('W', 'W', 'W', 'L'), n_points = 50) %>% 
  plot_posterior_grid()
```

```{r}
posterior_grid_approximate(prior_step, c('L', 'W', 'W', 'L', 'W', 'W', 'W'), n_points = 50) %>% 
  plot_posterior_grid()
```

### 2M3

#### Answer

$\mathrm{Pr(Earth|land)} = \dfrac{\mathrm{Pr(land|Earth)Pr(Earth)}}{\mathrm{Pr(land)}} = 0.3 * 0.5 / (0.3 * 0.5 + 1 * 0.5) = 0.23$

### 2M4

#### Answer

The possibilities for the chosen card and orientation are:

* Card 1: BB (Face-up colour, Face-down colour)
* Card 1: BB
* Card 2: BW
* Card 2: WB
* Card 3: WW
* Card 3: WW

There are 3 scenarios where B is the face-up colour, and in 2 of those B is also the face-down colour. Since each scenario is equally likely, the probability is $2/3$.

### 2M5

#### Answer

We add the following scenarios to the explanation from 2M4:

* Card 4: BB
* Card 4: BB

Now there are 5 scenarios where B is the face-up colour, and in 4 of those B is also the face-down colour. Since each scenario is equally likely, the probability is $4/5$.

### 2M6

#### Answer

We weight the scenarios in 2M4 by the probability that each card is pulled. The weights are (1, 2, 3) for Cards 1, 2, and 3 respectively, so we "double" the weight of the scenario `Card 2: BW`. This gives 4 scenarios where black is face-up, and in 2 of those black is the other side. So the probability is $1/2$.

### 2M7

#### Answer

We fan out the three scenarios that have black showing face-up:

* Card 1: BB -> Card 2: W
* Card 1: BB -> Card 3: W
* Card 1: BB -> Card 2: W
* Card 1: BB -> Card 3: W
* Card 2: BW -> Card 3: W

Note that in the last scenario the new card must be Card 3. Now the probability that black is showing face-down on the first card is $4/5 = 0.8$.

### 2H1

#### Answer

We want

$\mathrm{Pr(Twins2|Twins1)} = \dfrac{\mathrm{Pr(Twins1, Twins2)}}{\mathrm{Pr(Twins1)}}$

$\mathrm{Pr(Twins1)} = 0.5 * 0.1 + 0.5 * 0.2 = 0.15$

$\mathrm{Pr(Twins1, Twins2)} = 0.5 * (0.1^2)  + 0.5 * (0.2^2) = 0.025$

So the probability of the next birth being twins is $0.025 / 0.15 = 1/6$

### 2H2

#### Answer

$\mathrm{Pr(SpeciesA|Twins1)} = \dfrac{\mathrm{Pr(Twins1|SpeciesA)Pr(SpeciesA)}}{\mathrm{Pr(Twins1)}}$

$\mathrm{Pr(Twins1)} = 0.15$ from above, and $\mathrm{Pr(SpeciesA)} = 0.5$, and $\mathrm{Pr(Twins1|SpeciesA)} = 0.1$

So the probability is $0.1*0.5 / 0.15 = 1/3$

### 2H3

#### Answer

Let `T` denote a twin birth and `S` denote a single birth. We want

$\mathrm{Pr(SpeciesA|[TS])} = \dfrac{\mathrm{Pr([TS]|SpeciesA)Pr(SpeciesA)}}{\mathrm{Pr([TS])}}$

$\mathrm{Pr([TS]|A)} = 0.1 * 0.9 = 0.09$

$\mathrm{Pr([TS])} = 0.5 * 0.1 * 0.9 + 0.5 * 0.2 * 0.8 = 0.125$

The the probability that the panda is from species A is $0.09 * 0.5 / 0.125 = 0.36$

### 2H4

#### Answer

We want $\mathrm{Pr(SpeciesA|TestA)} = \dfrac{\mathrm{Pr(TestA|SpeciesA)Pr(SpeciesA)}}{\mathrm{Pr(TestA)}}$.

**Without the birth information**, $\mathrm{Pr(SpeciesA)} = 0.5$.

$\mathrm{Pr(TestA)}$ is the probability that the test returns positive for species A. This could happen if the panda is actually species A with 0.8 probability, or if the panda is species B with 0.35 probability. Then $\mathrm{Pr(TestA)} = 0.5 * 0.8 + 0.5 * 0.35 = 0.575$.

So the posterior probability that it is species A is $0.8 * 0.5 / 0.575 = 0.23$

**With the birth information**, we would believe $\mathrm{Pr(SpeciesA)} = 0.36$. So we change the weights for Species A, B in our calculations from (0.5, 0.5) to (0.36, 0.64).

Then the posterior probability that the panda is species A is $0.8 * 0.36 / (0.8 * 0.36 + 0.64 * 0.35) = 0.5625$.

Intuitively, since the birth data showed a twin birth and twin births are more likely for species B, we are less confident that the panda is species A when we incorporate this information.
